---
title: "FRED API Data Vintage Example"
author: "Lars Vilhuber"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document demonstrates the importance of using vintage dates when pulling API data from FRED.

## Load Required Libraries

```{r load-libraries, message=FALSE, warning=FALSE}
if (!require("fredr")) install.packages("fredr")
if (!require("dplyr")) install.packages("dplyr")
library(fredr)
library(dplyr)
```

## Set Parameters

```{r parameters}
# Set some parameters we want to re-use
# - the date range we want
DATE_START <- as.Date("2007-01-01")
DATE_END <- as.Date("2007-01-01")
# - the as-of date we want
CURRENT_DATE <- Sys.Date()
VINTAGE <- as.Date("2021-12-31")
# - for testing, other as-of dates
ALTVINTAGES <- as.Date(c("2008-09-15", "2015-09-15"))
NOTE <- "README ::::"
```

## Pulling Data WITHOUT Vintage Specification

By default, we pull data without specifying a vintage:

```{r pull-default, echo=FALSE}
data_default <- fredr(
  series_id = "GNPCA",
  observation_start = DATE_START,
  observation_end = DATE_END
)
print(data_default)
default_value <- mean(data_default$value, na.rm = TRUE)
```

## Pulling Data WITH Vintage Specification


Now we do the same thing, but precisely defining the vintages:

```{r pull-vintaged, echo=FALSE}
data_vintaged <- fredr(
  series_id = "GNPCA",
  observation_start = DATE_START,
  observation_end = DATE_END,
  vintage_dates = VINTAGE
)
print(data_vintaged)
vintaged_value <- mean(data_vintaged$value, na.rm = TRUE)
```

## Compare the Results

As of `r CURRENT_DATE`, the two values are:

- `r default_value`,  when not specifying a vintage
- `r vintaged_value`,  when specifying vintage `r VINTAGE`

**Expected result:** As of the current date, the two values may differ because:

- The default pulls the most recent revision
- The vintaged data pulls the value as it was known on 2021-12-31

## Pulling Multiple Vintages for Comparison


Now let's see why this matters - let's pull down a few more vintages:

```{r multiple-vintages, echo=FALSE}
# Combine all vintage dates
all_vintages <- c(VINTAGE, ALTVINTAGES)

# Pull data for each vintage
vintage_data_list <- lapply(all_vintages, function(vintage) {
  data <- fredr(
    series_id = "GNPCA",
    observation_start = DATE_START,
    observation_end = DATE_END,
    vintage_dates = vintage
  )
  data$vintage <- as.character(vintage)
  return(data)
})

# Combine all vintage data
vintage_comparison <- do.call(rbind, vintage_data_list)

# Reshape for comparison
vintage_wide <- vintage_comparison %>%
  select(date, value, vintage) %>%
  tidyr::pivot_wider(names_from = vintage, values_from = value, names_prefix = "vintage_")
```


## Comparison of GNPCA values across different vintages"

```{r print_vintages}
print(vintage_wide)
```


*NOTE: These values should NEVER change once recorded. The value for 2007-01-01 as of 2008-09-15 will always be what it was on that date.


## Key Lessons

**LESSON:**

1. Always use a fixed vintage date to query the API for reproducibility
2. Some series change as time progresses, even for historical values
3. Without vintage specification, you get the latest revision

## Data Persistence Strategy

**SUPPLEMENTARY LESSON:** Save the data pulled through the API as an intermediate dataset and if permissible by the license (check!), redistribute it in case that the API is deprecated and won't work in the future.

### Data Persistence Strategy

```{r data-persistence-code, echo=FALSE}
# Create directories if they don't exist
dir.create("data", showWarnings = FALSE)
dir.create("data/fred", showWarnings = FALSE)

# Check if file exists
if (file.exists("data/fred/fred_gnpca.rds")) {
  cat(NOTE, "Re-using existing file\n")
  fred_data <- readRDS("data/fred/fred_gnpca.rds")
} else {
  # Code if the file does not exist
  # You could do the full API pull
  # conditional on the intermediate
  # file NOT being there
  cat(NOTE, "Reading in data from FRED API with vintage =", as.character(VINTAGE), "\n")
  fred_data <- fredr(
    series_id = "GNPCA",
    observation_start = DATE_START,
    observation_end = DATE_END,
    vintage_dates = VINTAGE
  )
  saveRDS(fred_data, "data/fred/fred_gnpca.rds")
}
```

### Final Dataset

```{r data-persistence-print, echo=FALSE}
print(fred_data)
```

### Analysis Complete

Intermediate data saved to: `data/fred/fred_gnpca.rds`

This ensures reproducibility even if the API changes or becomes unavailable.

## Summary

This analysis demonstrates the critical importance of specifying vintage dates when working with FRED API data to ensure reproducibility and understand how economic data revisions affect historical values.
